# Chi squared test

The chi-square test provides a test for the __goodness of fit__, where _fit_ means how closely the observed results agree with the expected results.

The formula for the chi-square test statistic is:

## Chi-squared probability densities 

We can determine the probability density for the chi square distribution for 1-4 degrees of freedom using the `dchisq()` function

```{r}
x <- 0:20
curve(dchisq(x,1),0,20, ylab="density", xlab="chi square", lwd=3, ylim=c(0,0.9))
curve(dchisq(x,2),0,20, add=T, col=2, lwd=3)
curve(dchisq(x,3),0,20, add=T, col=3, lwd=3)
curve(dchisq(x,4),0,20, add=T, col=4, lwd=3)
```

## Chi-square cumulative probability distributions 

We can determine the cumulative probability distributions for the chi square distribution for for 1-4 degrees of freedom using the `pchisq()` function

```{r}
x <- 0:20
curve(pchisq(x,1),0,20, ylab="p", xlab="chi square", lwd=3)
curve(pchisq(x,2),0,20, add=T, col=2, lwd=3)
curve(pchisq(x,3),0,20, add=T, col=3, lwd=3)
curve(pchisq(x,4),0,20, add=T, col=4, lwd=3)
```

## Assess results of mendel's dihybrid cross using chi-square test statistic

```{r}
# probability density for 3 degrees of freedom 
x <- 0:50
curve(dchisq(x,3),0,20, ylab="density", xlab="chi square", lwd=3, ylim=c(0,0.3), col=3)

#probability density for 3 degrees of freedom; density for chisq > 0.47 indicated using polygon() 
x <- 0:50
curve(dchisq(x,3),0,20, ylab="density", xlab="chi square", lwd=3, ylim=c(0,0.3), col=3)
xvals <- seq(0.47,20,length=100)
dvals <- dchisq(xvals,3)
polygon(c(xvals,rev(xvals)),c(rep(0,100),rev(dvals)), col=3)

#cumulative distribution for 3 degrees of freedom with chisq=0.47 indicated using abline()
x <- 0:20
curve(pchisq(x,3),0,20, ylab="p", xlab="chi square", lwd=3, col=3)
abline(v=0.47)
abline(h=pchisq(0.47,3))
```

## Assessing the results from the Bateson and Punnetts's dihybrid cross using chi-square test statistic

```{r}
#probability density for 1 degrees of freedom 
x <- 0:50
curve(dchisq(x,1),0,50, ylab="density", xlab="chi square", lwd=3, ylim=c(0,0.5), col=1)

#probability density for 1 degrees of freedom; density for chisq > 3395.552 indicated using polygon() 

```

### Log transformation

The problem is that the relevant chi square value is a very large number
one. We can do this using the `log10()` function

```{r}
x <- 0:4000
curve(dchisq(x,1),0,log10(4000), ylab="density", xlab="log10(chi square)", lwd=3, ylim=c(0,2.2), col=1)
```

Now let's look at the probability denisty from 3395.52 in log10 space

```{r}
x <- 0:4000
curve(dchisq(x,1),0,log10(4000), ylab="density", xlab="log10(chi square)", lwd=3, ylim=c(0,2.2), col=1)
xvals <- seq(log10(3395.552),log10(4000),length=100)
dvals <- dchisq(xvals,1)
polygon(c(xvals,rev(xvals)),c(rep(0,100),rev(dvals)), col=1)
```

similarly, we can look at the cumulative distribution for 1 degrees of freedom with chisq=3395.552 indicated using `abline()`

```{r}
x <- 0:4000
curve(pchisq(x,1),0,log10(4000), ylab="p", xlab="log10(chi square)", lwd=3, col=1)
abline(v=log10(3395.552))
abline(h=pchisq(log10(3395.552),1))
```

## Analysis of Bateson and Punnett's data using internal R functions

Bateson and Punnett's data.  Note that data are entered COLUMN-WISE for a matrix.

```{r}
bp <- matrix(c(4831,390,393,1338),nrow=2,
	dimnames=list(c("long","round"),c("purple","red"))
)
```

Perform chi-square test using the R function `chisq.test()`

```{r}
chisq.test(bp, correct=F)
```

and, visualize the contingency table

```{r}
mosaicplot(bp, cex.axis=2, main="", color=T)
```










