# Nonparametric methods {#nonparametric}

## The sign test.

The non-parametric alternative to the one sample or paired t-test 

```{r}
before <- c(5,100,2000,20,1000)
after <- c(8,98,2200,16,900)

#a paired t-test returns the following result
t.test(before,after, paired=T, equal.var=T)

#We can implement a sign test ourselves by:
#1. calculate the difference 
diff <- after-before
#2 count how many are +ve (greater than zero)
sum(diff>0) #there are two + values of 5 total
#Compute the probability of getting two or less values greater than zero from 5 "trials" (one side) or two or less values less than zero from 5 "trials" if the null hypothesis is true (i.e. eqal probability of getting + or - ; p=0.5).
binom.test(2,5,0.5, alternative="two.sided") #p-value=1 
```

## Performing sign tests in R

Alternatively, the wilcox.test() function performs a paired sample test of the Wilcoxon signed rank test of the null that the distribution of x - y (in the paired two sample case) is symmetric about mu.

```{r}
wilcox.test(before,after, paired=T) #also returns a p-value of 1
```


## A more interesting example.  

Generating two random samples or 300 values fron uniform distributions that differ slightly

```{r}
set.seed(1)
before1 <- runif(300, min=-0.5,max=0.5)
set.seed(2)
after1 <- runif(300,min=-0.45,max=0.55)

diff1 <- after1-before1
sum(diff1<0) #number of values out of 300 that are below zero (-ve)
binom.test(sum(diff1<0),300,0.5, alternative="two.sided") #binomial test 


wilcox.test(before1,after1, paired=T) #p-value is not 1
t.test(before1,after1, paired=T) #compare result with the parametric t test.

```

## The Mann-Whitney test.

An alternative to two sample t-test

```{r}

sample1 <- c(1.5,5,20,30)
sample2 <- c(2.5,4,19,29,32)

wilcox.test(sample2,sample1)
```


A more interesting example.  generating two random samples or 300 values fron uniform distributions that differ slightly

```{r}
set.seed(3)
sample3 <- runif(300, min=-5,max=5)
set.seed(4)
sample4 <- runif(300,min=-4.5,max=5.5)

#the samples are not very normal
hist(sample3)
hist(sample4)

#let's check the means
mean(sample3)
mean(sample4)

wilcox.test(sample3,sample4, paired=F) #p-value is not 1
t.test(sample3,sample4) #compare result with the parametric t test
```

## Spearman (rank) correlation

```{r}
tumor.grade <- c(1,2,2,3,3,4,5,5)
gene.expression <- c(20,20,30,40,50,40,40,50)

cor.test(tumor.grade,gene.expression) #pearson linear correlation by default

cor.test(tumor.grade,gene.expression, method="spearman") #pearson linear correlation by default
```

### Comparing spearman and pearson correlation 

Will often look similar,  but for the example below they differ

```{r}

data1 <- 1:20
data2 <- 1/data1
plot(data1,data2)

cor.test(data1,data2,method="pearson") # = -0.707623
cor.test(data1,data2,method="spearman") # = -1
```

